Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities

### 2017-05-22 paper.Rmd
big update

bibliography: library.bib
output: rticles::elsevier_article
---

@@

Mozabique's recent economic growth has been facilitated by the government's open policy towards foreign investment, a plentiful supply of natural resources (particularly minerals and hydrocarbons) [@Rogers] and relatively inexpensive labor. Improvements in health have accompanied economic expansion, but Mozambique still lags behind in basic health outcomes - particularly those related to malaria - even by regional standards. Despite  strong evidence showing that improved health yields substantial economic benefits [@World1999, @Bloom2008], direct foreign investment (FDI) under the auspices of "corporate social responsibility" (CSR) has largely ignored malaria prevention in favor of other social good activities.

The few malaria-related CSR projects in recent years has come in large, "mega-projects" whose well-publicized malaria abatement activities are profit-driven [@Mouzin2011] or whose primary aim is not profit-related [@Han]. Though the former is often portrayed as a "win-win" for business and public health, the latter also offers tangible benefits for private industry, and should be understood as operating under the same conditions and with the same motivations.

Beginning in late 2015, the flood of FDI into Mozambique (and many other subsaharan African nations) slowed to a trickle. The impact of this slowdown on CSR is not yet known, but it can reasonably be assumed that it will mean a reduction in CSR activities (albeit with lag).  Given the rapidly changing economic and epidemiologic context in Mozambique, a comprehensive and current understanding of both (a) the landscape of FDI and CSR in Mozambique and (b) public health issues (namely, malaria control) which are directly affected those investments is sorely needed. Such an understanding may foster greater private interest in for-profit investment in public health measures. Likewise, it can facilitate better public sector understanding of potential industry partners and stakeholders (a prerequisite to greater public-private collaborations), and guide the public sector away from an inflexible dependence on FDI for the provision of public health necessities.

We carried out a systematic review of economic and public health research pertaining to FDI, CSR, malaria and Mozambique. This paper gives an overview of trends in FDI and CSR in Mozambique, with a focus on its impact on malaria control. It is by no means comprehensive, but offers a consolidated starting point for understanding both the place of FDI and CSR in the literature, as well as where the interests and incentives of the public and private sectors converge and differ in regards to malaria control.

## FDI and CSR in Mozambique

@@

# PLOT
g1 <- 
  ggplot(data = temp[temp$key %in% unique(sort(temp$key))[1:2],], 
         aes(x = year, y = value)) +
    geom_line() +
    facet_grid(.~key) +
    theme_bw()
g2 <- 
  ggplot(data = temp[temp$key %in% unique(sort(temp$key))[3:4],], 
         aes(x = year, y = value)) +
    geom_line() +
    facet_grid(.~key) +
    theme_bw()
multiplot(g1, g2)  
```

@@

# Systematic review of FDI and CSR in health research

## Search

if('search_results.RData' %in% dir()){
  load('search_results.RData')
} else {
  csr_results <- pubmed(start_year = 2000,
                         end_year = 2016,
                         search_topic = '(malaria) and (corporate social responsibility)',
                         counts_only = FALSE,
                         n_records = 5000)  
  fdi_results <- pubmed(start_year = 2000,
                       end_year = 2016,
                       search_topic = '(malaria) and (foreign direct investment)',
                       counts_only = FALSE,
                       n_records = 5000)  
    moz_csr_results <- pubmed(start_year = 2000,
                     end_year = 2016,
                     search_topic = '(mozambique) and (corporate social responsibility)',
                     counts_only = FALSE,
                     n_records = 5000)  
  moz_fdi_results <- pubmed(start_year = 2000,
                     end_year = 2016,
                     search_topic = '(mozambique) and (foreign direct investment)',
                     counts_only = FALSE,
                     n_records = 5000)  
  save(csr_results,
       fdi_results,
       moz_csr_results,
       moz_fdi_results,
       file = 'search_results.RData')
}

1. `"(malaria) and (corporate social responsibility)"`
2. `"(malaria) and (foreign direct investment)"`
3. `"(mozambique) and (corporate social responsibility)"`
4. `"(mozambique) and (foreign direct investment)"`

@@

## Qualitative synthesis

@@

# Scaling up malaria control through CSR: opportunity and risk

>>>>

bibliography: library.bib
output: rticles::elsevier_article
header-includes:
     - \usepackage{longtable}
---

@@

Mozabique's recent economic growth has been facilitated by the government's open policy towards foreign investment, a plentiful supply of natural resources (particularly minerals and hydrocarbons) [@Rogers] and relatively inexpensive labor. Improvements in health have accompanied economic expansion, but Mozambique still lags behind in basic health outcomes - particularly those related to malaria - even by regional standards. Most economic activities have to deal
with malaria: particularly agricultural-based and extraction activities cannot be carried out without consistent investment towards malaria management, both
prevention and prompt treatment. This can be done as part of the “normal activities” of private sector or can go slightly beyond what is strictly needed for the
activities and be seen as part of the corporate social responsibilityDespite  strong evidence showing that improved health yields substantial economic benefits [@World1999, @Bloom2008], there is little evidence regarding the extent to which "corporate social responsibility" (CSR) activities of foreign firms have targeted or impacted malaria. Understanding the relationship between FDI, CSR and malaria is essential to identifying which incentives and barriers to private sector involvement in malaria elimination exist, as well as identifying potential opportunities and synergies for scaling up private sector involvment in malaria elimination.

That said, 
The few malaria-related CSR projects in recent years has come in large, "mega-projects" whose well-publicized malaria abatement activities are profit-driven [@Mouzin2011] or whose primary aim is not profit-related [@Han]. Though the former is often portrayed as a "win-win" for business and public health, the latter also offers tangible benefits for private industry, and should be understood as operating under the same conditions and with the same motivations.

Private foreign investors showed a great optimism during the FMI meeting held in Maputo in 2014 (reference needed). Beginning in late 2015, the flood of FDI into Mozambique (and many other subsaharan African nations) slowed to a trickle. The impact of this slowdown on CSR is not yet known, but it can reasonably be assumed that it will mean a reduction in CSR activities (albeit with lag). Almost in parallel to this, Mozambique has had serious financial problems which have
resulted in default (reference needed http://www.economist.com/news/middleeast-and-africa/21715030-mozambique-fails-pay-its-debtsmozambiques-default). Given the rapidly changing economic and epidemiologic context in Mozambique, a comprehensive and current understanding of both (a) the landscape of FDI and CSR in Mozambique and (b) public health issues (namely, malaria control) which are directly affected those investments is sorely needed. Such an understanding may foster greater private interest in for-profit investment in public health measures. Likewise, it can facilitate better public sector understanding of potential industry partners and stakeholders (a prerequisite to greater public-private collaborations), and guide the public sector away from an inflexible dependence on FDI for the provision of public health necessities.

We carried out a systematic review of economic and public health research pertaining to FDI, CSR, malaria and Mozambique. This paper gives an overview of trends in FDI and CSR in Mozambique as well as a synthesis of academic literature on the topic, with a focus on its impact on malaria control. It is by no means comprehensive, but offers a consolidated starting point for understanding both the place of FDI and CSR in the literature, as well as where the interests and incentives of the public and private sectors converge and differ in regards to malaria control.

# Methods

if('search_results.RData' %in% dir()){
  load('search_results.RData')
} else {
  csr_results <- pubmed(start_year = 2000,
                         end_year = 2016,
                         search_topic = '(malaria) and (corporate social responsibility)',
                         counts_only = FALSE,
                         n_records = 5000)  
  fdi_results <- pubmed(start_year = 2000,
                       end_year = 2016,
                       search_topic = '(malaria) and (foreign direct investment)',
                       counts_only = FALSE,
                       n_records = 5000)  
    moz_csr_results <- pubmed(start_year = 2000,
                     end_year = 2016,
                     search_topic = '(mozambique) and (corporate social responsibility)',
                     counts_only = FALSE,
                     n_records = 5000)  
  moz_fdi_results <- pubmed(start_year = 2000,
                     end_year = 2016,
                     search_topic = '(mozambique) and (foreign direct investment)',
                     counts_only = FALSE,
                     n_records = 5000)  
  save(csr_results,
       fdi_results,
       moz_csr_results,
       moz_fdi_results,
       file = 'search_results.RData')
}
pubmed_results <- rbind(csr_results$results %>% mutate(type = 'CSR'),
                        fdi_results$results %>% mutate(type = 'FDI'),
                        moz_csr_results$results %>% mutate(type = 'CSR MOZ'))

We sought to identify and describe sources of information regarding FDI and CSR in Mozambique, particularly in regards to malaria control and elimination. We carried out this identification process via 2 distinct approaches to understanding FDI, CSR and malaria in Mozambique:

1. Identification and analysis of quantitative datasets.
2. Systematic review of both grey literature (news, company websites, etc.) and academic literature.

## Quantitative datasets

We examined and analyzed trends from multiple datasets related to FDI, CSR and malaria in Mozambique. These included the "Doing Business" data pertaining to the measurement of regulatory efficiency, the GADM, the Deutsche Bundesbank Data Repository for foreign exchange rate history, the Knoema World Data Atlas for macroeconomic trends, the World Bank data portal for information containing to net foreign inflows and FDI, USAID Demographic and Health Survey data on sociodemographics and health-related practices, the Instituto Nacional de Estatistica for granular data pertaining to Mozambique's population and economic activities, and Institute for Health Metrics and Evaluation data pertaining to disease trends over time.

## Systematic review

### Grey literature

Our grey literature review followed known practices [@Godin2015, @Adams2016], relying on internet searches and linked references. Sources include, but are not limited to UN reports on economic trends, WHO reports and brochures pertaining to health trends, company websites and brochures, newspaper articles, and think-tank white papers.

### Academic literature

1. `"(malaria) and (corporate social responsibility)"` (no results found in EBSCO)
2. `"(malaria) and (foreign direct investment)"`
3. `"(mozambique) and (corporate social responsibility)"` (no resuts found in EBSCO)
4. `"(mozambique) and (foreign direct investment)"`

# Results

## Quantitiative datasets pertaining to FDI and CSR in Mozambique

@@

# PLOT
g1 <- 
  ggplot(data = temp[temp$key %in% unique(sort(temp$key))[1:2],], 
         aes(x = year, y = value)) +
    geom_line() +
    facet_grid(.~key) +
    theme_bw() +
  theme(strip.text = element_text(size=5))
g2 <- 
  ggplot(data = temp[temp$key %in% unique(sort(temp$key))[3:4],], 
         aes(x = year, y = value)) +
    geom_line() +
    facet_grid(.~key) +
    theme_bw() +
   theme(strip.text = element_text(size=5))

multiplot(g1, g2)  
```

@@

## Systematic review

### Grey literature

#### Grey literature search strategy

We devised 2 simple search queries, and use www.google.com and www.bing.com to retrieve results. The queries were:

1. `Mozambique foreign direct investment malaria`
2. `Mozambique corporate social responsibility malaria`

#### Grey literature search results

Google returned 266,000 results for the former query, and 240,000 for the latter (506,000 in total); Bing returned 4,620 and 20,900, respectively. We screened the top 50 results from both services (a total of 100 items) for both queries For the former, of the 100 items identified, 33 were identical in both search engines (yielding a total of 67 unique items); in the latter, 39 were identical (yielding a total of 61 unique items).

\begin{center}
\includegraphics[width=400pt]{img/prisma_grey.png}
\end{center}

#### Grey literature synthesis

Our grey literature review returned relatively little information which was not readily available in those sources uncovered in the data review and systematic review. Very few foreign companies operating in Mozambique have any publicly available information regarding corporate social responsibility activities.

One notable exception is the Nando's-lead Goodbye Malaria Trust, an umbrella organization which inclues a development impact bond, the Goodbye Malaria initiative, and partnerships with several other foreign firms operating in Mozambique (more details here).

### Academic literature

@@

The below table shows the results retrieved for the systematic review. Full article information is in the bibliography

```{r}
search_results <-
  pubmed_results %>%
  mutate(source = 'Pubmed') %>%
  dplyr::select(source, year, title, type, `last _author_last_name`) %>%
  rename(author = `last _author_last_name`)

# Manually add in Elisa's search results
econ_results <-
  data_frame(source = rep('EBSCOhost', 20),
             year = NA,
             title = NA,
             type = NA,
             author = NA)
econ_results[1,2:5] <- c(2014, 
                         'A Critical Review of FDI Inflows and Economic Growth in Low-Income SADC Countries: Prospects and Challenges',
                         'FDI MOZ',
                         'Mahembe')
econ_results[2,2:5] <- c(2000, 'Administrative barriers to foreign investment: Reducing red tape in Africa', 'FDI MOZ', 'Emery')
econ_results[3,2:5] <- c(2012,
                         'Attracting Foreign Direct Investment in Developing Countries: Determinants and Policies-A Comparative Study between Mozambique and China',
                         'FDI MOZ', 'Tembe')
econ_results[4,2:5] <- c(2013,
                         'Contemporary Processes of Large-Scale Land Acquisition in Sub-Saharan Africa: Legal Deficiency or Elite Capture of the Rule of Law?',
                         'FDI MOZ', 'German')
econ_results[5,2:5] <- c(2012,
                         'Corruption and Multinational Companies Entry Modes--Do Linguistic and Historical Ties Matter?',
                         'FDI MOZ', 'Grande')
econ_results[6,2:5] <- c(2014,
                         'Determining the Nature and Extent of Spillovers: Empirical Assessment',
                         'FDI MOZ', 'Winkler')
econ_results[7,2:5] <- c(1999, 'Foreign Direct Investment in Africa: Policies Also Matter',
                         'FDI MOZ', 'Morisset')
econ_results[8,2:5] <- c(2002,
                         'Foreign Direct Investment in Africa--Some Case Studies',
                         'FDI MOZ', 'Bassu')
econ_results[9,2:5] <- c(2014,
                         'Growth, Capital Accumulation and Economic Porosity in Mozambique: Social Losses, Private Gains',
                         'FDI MOZ', 'Castel-Branco')
econ_results[10,2:5] <- c(2007,
                          'Linkage Between foreign Direct Investment, Trade and Trade Policy: An Economic Analysis with Application to the Food Sector in OECD Countries and Case Studies in Ghana, Mozambique, Tunisia and Uganda',
                          'FDI MOZ', 
                          'Wilson')
econ_results[11,2:5] <- c(2012,
                          'Mining FDI and Infrastructure Development on Africas East Coast: Examining the Recent Experience of Tanzania and Mozambique',
                          'FDI MOZ',
                          'Robbins')
econ_results[12,2:5] <- c(2013,
                          'Potential and actual FDI spillovers in global value chains: the role of foreign investor characteristics, absorptive capacity and transmission channels',
                          'FDI MOZ',
                          'Winkler')
econ_results[13,2:5] <- c(2004,
                          'Regional integration, FDI and competitiveness in southern Africa',
                          'FDI MOZ',
                          'Goldstein')
econ_results[14,2:5] <- c(2014,
                          'Sector Case Study: Mining',
                          'FDI MOZ',
                          'Barnard')
econ_results[15,2:5] <- c(2011,
                          'Strategic Privatisation: Rehabilitating the Mozambican Sugar Industry',
                          'FDI MOZ',
                          'Buur')
econ_results[16,2:5] <- c(2016,
                          'The Economics and Politics of Local Content in African Extractives: Lessons from Tanzania, Uganda and Mozambique',
                          'FDI MOZ',
                          'Hansen')
econ_results[17,2:5] <- c(2002,
                          'The Role of FDI in Economic Development',
                          'FDI MOZ',
                          'Bjorvatn')
econ_results[18,2:5] <- c(2010,
                          'Uncovering Trends in the Accumulation of Technological Capabilities and Skills in the Mozambican Manufacturing Sector',
                          'FDI MOZ',
                          'Warren-Rodriguez')
econ_results[19,2:5] <- c(2009,
                          'Public Governance, Health and Foreign Direct Investment in Sub-Saharan Africa',
                          'FDI Malaria',
                          'Azemar')
econ_results[20,2:5] <- c(2014,
                          'Sector Case Study: Agribusiness',
                          'FDI MOZ',
                          'Barnard')

# Combine all results
econ_results$year <- as.numeric(as.character(econ_results$year))
search_results <-
  bind_rows(search_results,
            econ_results)

# Clean up
names(search_results) <- Hmisc::capitalize(names(search_results))
search_results$Author <- paste0(search_results$Author, ', et al')

# Make title shorter
add_n <- function(x, n = 20){
  nc <- nchar(x)
  if(nc < (n+15)){
    return(x)
  } else {
    starts <- seq(1, nc, n)
  stops <- c(seq(n, nc, n), nc)
  out <- list()
  for (i in 1:length(starts)){
    while(stops[i] <= nc &
          substr(x, stops[i], stops[i]) != ' '){
      stops <- stops + 1
    }
    out[[i]] <- substr(x = x, start = starts[i], stop = stops[i])
        while(!is.na(starts[i+1]) &
              starts[i + 1] <= stops[i]){
      starts[i + 1] <- starts[i + 1] +1
    }
  }
  out <- out[!out == '']
  final <- ''
  for (i in 1:(length(out) -1)){
    final <- paste0(final, trimws(out[[i]], 'both'), '\n', collapse = '')
  }
  final <- paste0(final, out[[length(out)]])
  return(final)

  }
}

# Make short funtion
make_short <- function(x, n = 20){
  if(nchar(x) < n){
    return(x)
  } else {
  return(paste0(substr(x, 1, n), '...', collapse = ''))
  }
}
for (s in 1:nrow(search_results)){
  search_results$Title[s] <- 
    make_short(search_results$Title[s])
    # add_n(search_results$Title[s], n = 60)
}
search_results$Year <- as.character(search_results$Year)
search_results <- search_results %>% arrange(Source, Type)
```

```{r, results = 'asis'}
# kable(search_results)
options(xtable.comment = FALSE)
print(xtable(search_results),
      size="\\fontsize{6pt}{7pt}\\selectfont",
      include.rownames = FALSE,
      include.colnames = TRUE,
      caption.placement = "top",
      comment=FALSE,
      floating=FALSE,
      tabular.environment='longtable')
```

The below chart shows our systematically retried publications by date. What is most striking is how little academic attention has been given to CSR, with none given to the intersection of CSR and malaria.

```{r}
x <- search_results %>%
  group_by(Year = as.numeric(Year), Type) %>%
  tally

cols <- colorRampPalette(brewer.pal(n = 9, 'Spectral'))(length(unique(x$Type)))
ggplot(data = x,
       aes(x = Year,
           y = n)) +
  geom_bar(color = 'darkgrey',
           stat = 'identity',
           position = 'stack',
           aes(#group = Type,
                 fill = Type)) +
  scale_fill_manual(name = 'Type', values = cols) +
  theme_bw() +
  labs(y = 'Articles',
       title = 'Systematic review articles by publication date')
```

### Qualitative synthesis of systematic review of academic literature

@@

# Discussion

## Scaling up malaria control through CSR: opportunity and risk
